---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# clinicalsignificance

<!-- badges: start -->

<!-- badges: end -->

The goal of clinicalsignificance is to conduct analyses of clinical significance in clinical intervention studies. In contrast to *statistical* significance, which assesses if it is probable that there is a treatment effect, *clinical* significance can be used to determine if a treatment effect is of practical use or meaninful for patients.

## Installation

You can install the development version of clinicalsignificance from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("pedscience/clinicalsignificance")
```

## Rationale

## Example
Given a tidy dataset, the employed instrument's reliability and descriptives (_M_ and _SD_) of the functional population, the clinical significance in a study can be easily assessed.

```{r example}
library(clinicalsignificance)

results <- claus_2020 %>% 
  clinical_significance(
    id = id, 
    time = time, 
    outcome = bdi, 
    pre = 1, 
    post = 4, 
    reliability = 0.81, 
    m_functional = 8, 
    sd_functional = 8, 
    type = "c"
  )

results
```

You can receive a detailed summary of the analysis by

```{r}
summary(results)
```

or plot the results with

```{r}
plot(results)
```

