---
title: "Anchor-Based Approaches"
date: "September 26, 2023"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Anchor-Based Approaches}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
In this beginner-friendly tutorial, we will explore anchor-based approaches to clinical significance using R. Anchor-based approaches are commonly used in clinical research to assess the meaningfulness of change in patient outcomes. We will be working with the `claus_2020` dataset and the `cs_anchor()` function to demonstrate various aspects of these approaches.

## Prerequisites
Before we begin, ensure that you have the following prerequisites in place:

- R installed on your computer.
- Basic understanding of R programming concepts.

## Looking at the Datasets
First, let's have a look at the datasets, which come with the package.

```{r}
library(clinicalsignificance)

antidepressants
claus_2020
```


# Individual Level Anchor-Based Approach
The individual level anchor approach is a method to determine the clinical significance of changes in individual patient outcomes over time. We will use the `cs_anchor()` function for this analysis.

## Basic Analysis
Let's start with a basic clinical significance distribution analysis using the `antidepressants` dataset. We are interested in the Mind over Mood Depression Inventory (`mom_di`) measurements and want to set Minimally Important Difference (MID) for an improvement to a value of 7.
```{r}
anchor_results <- antidepressants |> 
  cs_anchor(
    id = patient,
    time = measurement,
    outcome = mom_di,
    mid_improvement = 7
  )
```

Here's a breakdown of the code:

- `patient`, `measurement`, and `mom_di` are variables representing patient identifiers, assessment time points, and the Mind over Mood Depression Intentory (MoM-DI) scores, respectively.
- `mid_improvement` sets the  for improvement to 7.

## Handling Warnings
Sometimes, you may encounter warnings when using this function. You can turn off the warning by explicitly specifying the pre-measurement time point using the pre parameter. This can be helpful when your data lacks clear pre-post measurement labels.

```{r}
anchor_results <- antidepressants |> 
  cs_anchor(
    id = patient,
    time = measurement,
    outcome = mom_di,
    pre = "Before",
    mid_improvement = 7
  )
```


## Printing and Summarizing the Results
```{r}
# Print the results
anchor_results

# Get a summary
summary(anchor_results)
```

## Visualizing the Results
```{r}
# Plot the results
plot(anchor_results)

# Show clinical significance categories
plot(anchor_results, show = category)
```


## Data with More Than Two Measurements
When working with data that has more than two measurements, you must explicitly define the pre and post measurement time points using the pre and post parameters.

```{r}
claus_results <- claus_2020 |> 
  cs_anchor(
    id = id,
    time = time,
    outcome = bdi,
    pre = 1,
    post = 4,
    mid_improvement = 7
  )

summary(claus_results)
plot(claus_results)
```


## Group the Analysis
It is also possible to provide a grouping variable present in your data to group the analysis. The resulting plot distinguishes the effects for the provided groups.
```{r}
anchor_results_grouped <- claus_2020 |> 
  cs_anchor(
    id = id,
    time = time,
    outcome = bdi,
    pre = 1,
    post = 4,
    mid_improvement = 7,
    group = treatment
  )

anchor_results_grouped

# And plot the groups
plot(anchor_results_grouped)
```


## Analyzing Positive Outcomes
In some cases, higher values of an outcome may be considered better. You can specify this using the `better_is` argument. Let's see an example with the WHO-5 score where higher values are considered better. Suppose the MID is 4 in this case.

```{r}
anchor_results_who <- claus_2020 |> 
  cs_anchor(
    id = id,
    time = time,
    outcome = who,
    pre = 1,
    post = 4,
    mid_improvement = 4,
    better_is = "higher"
  )

anchor_results_who

# And plot the groups
plot(anchor_results_who)
```


# Group Level Anchor-Based Approach
The group level anchor-based approach assesses clinical significance for **groups** of patients, often in the context of treatment comparisons. Let's explore this approach.

## Performing the Analysis
```{r}
anchor_results_group_level <- claus_2020 |> 
  cs_anchor(
    id = id,
    time = time,
    outcome = bdi,
    pre = 1,
    post = 4,
    mid_improvement = 7,
    target = "group"
  )
```

## Summarizing the Results
```{r}
summary(anchor_results_group_level)
```

## Analysis for Different Groups
```{r}
claus_2020 |> 
  cs_anchor(
    id = id,
    time = time,
    outcome = bdi,
    pre = 1,
    post = 4,
    mid_improvement = 7,
    target = "group",
    group = treatment
  )
```

## Comparing Groups
```{r}
claus_2020 |> 
  cs_anchor(
    id = id,
    time = time,
    outcome = bdi,
    post = 4,
    mid_improvement = 7,
    target = "group",
    group = treatment,
    effect = "between"
  )
```

# Conclusion

Anchor-based approaches are valuable tools in clinical research for assessing the clinical significance of changes in patient outcomes. In this tutorial, we've covered the individual and group-level anchor approaches, and you've learned how to perform these analyses using R. These techniques can help researchers and healthcare professionals make informed decisions about the effectiveness of treatments and interventions.
